---
title: "SidAndrewFinalProject"
author: "Andrew Kuang"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r}
library(tidyverse)
```


# Question 1. File Linkage Integrity

## a) Read the four CSV files into R, building four data frames with the names “Cases”, “Parents”, “Children” and “Payments”. Show the dimensions of these data frames. (You may find it useful to save these data frames as Rdata objects in a file using the save command. You can then recover them with the load command more quickly than reading the CSV file.)

```{r}

csvToDf = function(v) {
  return(as.data.frame(read.csv(v)))
  
}

cases <- csvToDf("cases.csv")
parents <- csvToDf("parents.csv")
children <- csvToDf("children.csv")
payments <- csvToDf("payments.csv")

```

## b) What is the distribution of the number of children attached to a case? Show an appropriate plot of the distribution, and mark the location of the average number in the plot.

```{r}

number <- children %>% select(CASE_NUM, ID) %>% group_by(CASE_NUM) %>% mutate(numberchildren = length(ID)) %>% ungroup(CASE_NUM) %>% summarize(mean(numberchildren))


children %>% select(CASE_NUM, ID)%>% group_by(CASE_NUM) %>% mutate(numberchildren = length(ID)) %>%
  ggplot(mapping = aes(x = numberchildren)) + geom_histogram(binwidth = 1) + geom_vline(xintercept = as.numeric(number), size = 5, color = "red")
  

```
## c) The file children.csv may have more than one record for each child. What is the largest number of cases associated with a child, and indicate why you believe that this is indeed the same child.

```{r}

children %>% select(ID, CASE_NUM) %>% group_by(ID) %>% mutate(childCaseCount = n()) %>% tally() %>% summarize(max(n))
```

This is the same child because we grouped by ID, which identifies the child. Timmy has 12 cases of child support, as a child needing support. 

## d) Does every absent parent (AP_ID) identified in the payments data have an identifying record in the parents data file?

```{r}
# Gets all the unique AP_IDs from the payments data file
payments_AP_ID <- payments %>% select(AP_ID) %>% arrange() %>% pull()
payments_AP_ID <- unique(payments_AP_ID)

# Gets all the unique AP_IDs from the payments data file
parents_AP_ID <- parents %>% select(AP_ID) %>% arrange() %>% pull()
parents_AP_ID <- unique(parents_AP_ID)

# Checks if AP_IDs are identical
identical(payments_AP_ID, parents_AP_ID)
```

# Question 2: Recoding Categories

Some categorical variables among these data frames are sparse (seldom observed). For example, the variable PYMNT_SRC in Payments has category ‘M’ with 2 cases and category ‘R’ with 7. These are too few for modeling in regression.

Write a function named “pool_categories” that recodes a categorical variable into a “simpler” factor with fewer categories by pooling categories with counts below a threshold into a category labeled ‘Other’ (a factor level which your function should check does not already exist!). You might find the R function %in% useful for this exercise.

```{r}

payments.short <- payments %>% slice(1:1000)

pool_categories = function(variable, min){
  
  nameUse <- "Other";
  
  var.copy <- variable
  var_factor <- as.factor(var.copy)
  var_levels <- levels(var_factor)
  
  if(length(which(var_levels %in% "Other")) != 0)nameUse <- "Other1";
  
  levels(var_factor) <- c(levels(var_factor), nameUse)
  
  for(level in var_levels){
    if(sum(var.copy %in% level) < min){
      var.copy[which(variable == level)] <- nameUse
    }
  }
  return(as.factor(var.copy)) 
}

pool_categories(payments.short$PYMNT_SRC, 100)

```

# Question 3: Payment counts and amounts

You must use ggplot2 for generating the plots asked for in this question.

## (a) Make a variable Payments$DATE which is a viable R date by converting the COLLECTION_DT variable. Use this variable to find ### (i) the range of dates of all payments and 
```{r}
payments <- payments %>% mutate(DATE = as.Date(payments$COLLECTION_DT, "%m/%d/%Y")) %>% arrange(order(DATE))
payments[order(payments$DATE)]

payments[1,]
payments[length(payments),]
```


### (ii) the percentage of the total number of payments made before May 1, 2015.
## (b) Show a sequence plot of the total number of payments made on each day from May 1, 2015 through the end of the data.
## (c) What explains the bimodal shape of the marginal distribution of the number of payments over this time period? Explain with some evidence how you reached your opinion.
## (d) Describe the distribution of the payment amounts. Do you have an explanation for its shape? (You
might find it useful to work with a sample for plotting. R takes a while to draw 1.5 million points.)


```{r}

```


# Question 4: Payments for Cases

# Question 5: Consistency